The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-04-24 paper/ms.tex
new structure

\begin{figure*}
\begin{center}
  \includegraphics[width=0.8\textwidth]{f1.eps}
\end{center}
\caption{Shape of the \ly line for
    different maximum rotational velocities. The continuous (dashed)
    line represents the central (homogeneous) source distributions. 
    These profiles were obtained integrating the flux of photons 
	over all directions, this means we made histograms of the 
	numbers of photons escaping the galaxy with a given frequency.
    \label{fig:differentvelocities}}  
\end{figure*}

@@

@@

The central results of this paper are summarized in Figures
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. They show 2D histograms
of escape frequency $x$ and escape angle parameterized by
$\cos\theta$.

that rotation fix a preferred axis for both photons distributions, 
this implies that our systems doesn't have spherical symmetry and
we have to take into account the viewing angle $\theta$ for all
of our posterior analysis.

Figure \ref{fig:differentvelocities} shows the clear
impact of rotation on the morphology of the emergent \ly line.
This profiles were obtained integrating the flux of photons 
 over the angle $\theta = \pi/2$, this means we have made histograms of the 
 numbers of photons escaping the galaxy with a given frequency
The panel focus on the results for all of our symmetric models
 $\tau_{H}=(10^{5}, 10^{6}, 10^{7}$), showing that the
influence of rotation is present both when the photons are either
homogeneously or centrally initialized over the gas volume.

In the next subsections we characterize the line morphology by the
full width at half maximum (FWHM) and the location of the peak
maxima. In order to interpret the morphological changes in the line we
also report the median number of scatter for each \ly photon in the
simulation. For the models where dust is included we measure the 
escape fraction as a function of rotational velocity.

\subsection{Rotation and the spherical symmetry}

As a first approach we study the symmetry of the out-coming spectra
taking into account the line of view ($\theta$) of an observer. Figures 
\ref{fig:CentralSpec} and \ref{fig:HOMSpec} shows our findings, for 
both distributions the morphology of the line change as the line 
of sight of and observer changes. This is mainly due to the fact
that photons propagating at $\theta = \pi/2.0$ feels ... while photons in other direction feel
a less contribution.

We also consider possible variations in the flux with respect to $\theta$.
 We define the flux by:
\begin{equation}
F(\mu) = \frac{2\Delta N}{N\Delta \mu}, 
\end{equation}

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.4\textwidth]{muhisto.eps}
\end{center}
\caption{Position of the peak maxima as a function of rotational
  velocity $V_{\rm max}$. Continuous (dashed) lines correspond to
  central (homogeneous) source distributions. A value of $x_{\rm
    max}=0$ indicates that line becomes single
  peaked. \label{fig:muhisto}}  
\end{figure}

Taking into account these results we perform the subsequent analysis
for a fixed observer at $\theta = \pi/2.0$, the effect in the morphology
of the \lya	line for different velocities is clearly shown in Figure
 \ref{fig:differentvelocities}

\subsection{Line width and peak maxima}
\label{sec:widthpeak}

@@

There are three clear effects on the line's morphology as the
rotational velocity increases. First, the line
broadens; second, the double peaks reduce their intensity; and third,
the intensity at the line center rises, this effects are present for 
all the viewing angles. The last two effects merge to give the
impression that the double peaks are merger into one at high rotational
velocities, a result that is evident for the homogeneously distributed
sources as shown in the dashed lines of Figure \ref{fig:differentvelocities}.

@@

@@

\subsection{Average Number of Scatterings}

@@

\iffalse
\subsection{Anisotropy and off-centered emission}
\label{sec:off-center}

\begin{figure}
\begin{center}
  \includegraphics[width=0.40\textwidth]{f8.eps}
\end{center}
\caption{Schematic representation of the geometry for the off-center
  emission. The large circle represents the gas distribution and the
    two smaller circles the off-centered \ly emitting regions. The angular
    velocity vector lies along $\hat{k}$ direction
    towards the reader. The vector $\hat{k}_{\rm out}$ represents the
    outgoing direction of an escaping \ly photon and the observer is
    placed along the $\hat{j}$ direction. The distribution of $\Theta$
    angles is used to study potential emission anisotropy. 
    \label{fig:OCspheres}}
\end{figure}

\begin{figure}
  \includegraphics[width=0.42\textwidth]{f9.eps}
  \caption{Histogram of $\mu\equiv\cos\Theta$ for the 
    central (dotted), homogeneous (dashed) and off-center (solid
    line) distributions. Deviations from a flat distribution signal the presence of
    spatial emission anisotropy. 
    \label{fig:muhisto}} 
\end{figure}

\begin{figure*}
\begin{center}
\includegraphics[width=0.90\textwidth]{f10.eps}
\end{center}
\caption{\ly profiles for the off-centered emission. Dashed/solid
  lines represents spheres centered at $-0,5R/+0.5R$ respectively.
   \label{fig:OC_profiles}} 
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=0.90\textwidth]{f11.eps}
\end{center}
\caption{Triple peaked profiles for three different off-center positions
  $C=0.25R$ (doted line), $C=0.5R$ (dashed line) \& $C=0.75R$ (solid
  line) and for different rotational velocities. $\theta$ represents
  the position of the observer as mesured with the polar angle.
   \label{fig:3p_profiles}} 
\end{figure*}

Rotation breaks the spherical symmetry of the static model and
introduces the rotation axis as a preferred direction. This motivates
us to study the anisotropy of the outgoing photons. In order to study
possible deviations with respect to an isotropic flux we follow
\cite{Zheng2013} to estimate the flux as a function of the angle
$\Theta$ formed by the outgoing photons and the observer. Figure
\ref{fig:OCspheres} presents a geometric scheme to clarify the
definition of all relevant directions in the problem.

An estimator for the anisotropy is represented by
%
\begin{equation}
F(\mu) = \frac{2\Delta N}{N\Delta \mu}, 
\end{equation} 
%
where $\mu=\cos\Theta$, $N$ is the total number of outgoing photons,
$\Delta N$ is the number of photons in a angular bin $\Delta
\Theta$. This definition satisfies the condition
$\int_{-1}^{1}F(\mu)d\mu/2=1$.  With this definitions, the isotropic
emission is characterized by a flat $F(\mu)$ distribution.

Figure \ref{fig:OCspheres} presents the results for the $F(\mu)$
distribution for a sphere with $\tau_{\rm H}=10^{7}$ and $V_{\rm
  max}=200$\kms. The dashed and dotted lines represent the homogeneous
and central source distribution, respectively. This shows that the
 anisotropy induced by rotation is at the $3\%$ level. We
performed the same test with all the models and find that the
variation in $\mu$ does not depend on the rotational velocity;
however, high optical depth values can increase the variation of
$\mu$ up to a $15\%$.

We also explore a different source of anisotropy by studying the effect
of radiation sources located in off-centered spheres with respect to
the gas distribution. Figure \ref{fig:OCspheres} shows the two
spherical regions we define to test for this effect. Each has a radius
of $0.5R$, where $R$ is the radius of the gaseous sphere. With an
observer located along the $\hat{j}$ direction, we pick two possible
positions to center the spheres along the $\hat{i}$ axis: $\pm C\hat{i}$. In
this configuration the emitting sphere centered on $+C\hat{i}$ moves
towards the observer and the sphere centered on $-C\hat{j}$ moves
away from the observer.

In this case the $\mu$ distribution is not homogeneous as it is shown in
Figure \ref{fig:muhisto}. The off-center emission shows a variation of flux at
the levels of $10\% - 28\%$ in intensity with the angle $\Theta$. In general 
models with a larger $\tau_{H}$ present a higher variation in the flux.

We also explore the effects of the off-centered emission in the line
morphology. Figure \ref{fig:OC_profiles} show the spectra for the two
off-centered emitting spheres located at $C=\pm0.5R$.  In this case we
only use photons that have outgoing directions aligned with the
observer i.e. $0.9 <\vert \mu \vert < 1.0$. We do not find any strong
asymmetry in the line morphology within the range of explored velocities.

However, we find some cases where there is a possible signature of
triple peaked emission. This is only present in some models
($\tau_{\rm H} = 10^6, V=100$\kms and $\tau_{\rm H} = 10^7,
V=300$\kms) and could be the result of two competing effects: the rise
in the line's center as the photons in the photosphere escape in a
single excursion and the broadening in the two peaks as the photons
closer from the center have to scatter to escape. We examine in
greater detail these cases in the next sub-section.

\subsection{Triple Peaked Profiles}
\label{sec:3p}

The off-centered cases with hints of triple peaked emission motivate us
to explore this in more detail. The first condition is to increase the
number of photons in the radiative transfer calculation. In the line
profiles presented in the previous subsection, only $\sim 10^4$ photons
have the geometric conditions to build the line profile. We run a few
models with $N_{\gamma} = 10^6$ in order to improve the statistics and
explore smaller radii for the off-centered spheres.

We re-simulate four models that presented some signs of triple peaked
profiles: $V_{\rm max}=50\kms$, $\tau_{\rm H}=10^{5}$; $V_{\rm
  max}=100\kms$, $\tau_{H}=10^{6}$; $V_{\rm max}=150\kms$, $\tau_{\rm
  H}=10^6$; and $V_{\rm max}=200\kms$, $\tau_{\rm H}=10^{7}$. We
select the photons that were emitted within spheres of radius $0.25R$
centered at $C=(0.25R, 0.5R, 0.75R)$, following the same
geometry presented in Figure \ref{fig:OCspheres}.

Figure \ref{fig:3p_profiles} summarizes the results for all these
models. In all cases the stronger triple peaked signatures is
presented by the emitting spheres close centered at $0.75R$ close to
the surface. Emitting spheres closer to the center do not show clear
signatures for a triple peaked emission.

Furthermore, taking into account all the models that do not present hints for
a triple peaked emission it would seem that the rotational velocity
and optical depth must balance each other in order to show this feature.
 Namely, if the rotational velocity is close to half the FWHM
of the line, the triple peaks are more likely to occur. This is the
same threshold that in the case of centered emission marks the
difference between double and single peaked lines.

Are these features observable with current instrumental capabilities?
As an order of magnitude estimate we can consider that the width of
each one of the triple peaks is close to $1/3$ of the FWHM of the line
at a given $V_{\rm max}$. For the cases we have studied in this
subsection those values would be close to $50\kms$, $100\kms$ and
$200\kms$ for optical depths of $10^5$, $10^6$, and $10^7$
respectively. At a redshift of $z\sim 2-3$ each one of these widths
is multiplied by factors of $z+1\sim 3-4$. Considering a nominal value
for the resolution of $\Delta v=200\kms$ for a Near-IR spectrograph capable of observing
those galaxies \citep{Kulas12}, we find that only the case of triple
peaked emission for the most massive object $\tau_{\rm H}=10^{7}$,
$V_{\rm max}=200\kms$ would have a chance to be spectrally
resolved. We note that observations of triple peaked systems with
higher resolution spectrographs have been reported  by
\cite{Chonis2013}, however the large velocity off-sets ($\sim
1000$\kms) from the systemic redshift favor the hypothesis of an
interacting system, including emission from shocked and fluorecing
gas, as the origin of the triple peaked emission.
\fi

@@

Gas bulk rotation has a noticeable effect on the morphology of the
 \ly line. In this section we discuss the implications of these
 findings for the interpretation of available observational data. Our
 discussion focuses on the qualitative aspects of diverse
 morphological features observed by \cite{Kulas12} and \cite{Yamada2012}

We follow a classification of \ly profiles made by
\cite{Kulas12}. They presented observational results for star forming
galaxies at $z=2$-$3$ with multiple peaks in their \ly emission.
They classified the observed lines into five groups. In Group I the
red peak is stronger than the blue, while in Group II the blue peak is
stronger, in both cases the peaks are symmetrically located around the
line center. In Group III the blue peak also dominates but there is an
overall redward shift of the zero point. Group IV presents two similar
peaks symmetrically located around the zero point and Group V galaxies show
three peaks.

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; this could indicate the presence of rotation 
and \ly photons leaking though low column density holes, as in \cite{Behrens2014}. The same argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be reproduced by rotation with off-centered
emission given the right amount of asymmetry between the emitting
region and neutral hydrogen distribution.

Another relevant sample of observations was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Ahn2014,Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-center emission models with rotation Figure \ref{fig:3p_profiles}.

@@

\item For off-centered emission triple peaked profiles can emerge
  under two joint conditions: the emission occurs close to the sphere's
  surface and the rotational velocity is close to half the FWHM in the
  static case.   
\end{itemize}

Comparing our results with recent observed LAEs we find that many 
morphological features such as high central line flux, single peak
profiles and multiple-peaked profiles could be explained by gas bulk 
rotation present in these LAEs. We defer the study of triple peaked
emission in the context of gas bulk rotation to a future publication
(Remolina-Gutierrez et al., in preparation).

>>>>

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f1.eps}
\end{center}
\caption{Shape of the \ly line for different maximum rotational
  velocities for a viewing angle perpendicular to the rotation axis
  ($|\mu|\sim 0$). The continuous (dashed) line represents the central
  (homogeneous) source distributions.  The panels follow the same
  correspondence as Figures \ref{fig:CentralSpec} and \ref{fig:HomSpec}.
    \label{fig:differentvelocities}}  
\end{figure*}

@@

The central results of this paper are evident in Figures
\ref{fig:CentralSpec} and \ref{fig:HomSpec}. They show 2D histograms
of escape frequency $x$ and escape angle $\theta$ parameterized by
$|\mu|$ after verifying that the solutions are symmetric with respect to
$\mu=0$.  Making a selection around a given value of $\mu$ gives us the
line detected by an observer located at an angle $\theta$ with respect
to the rotation axis.

From these Figures we can see that the line properties change with
rotational velocity and depend on the viewing angle $\theta$.  In the
next subsections we describe in  detail the changes of the morphology
with velocity, optical depth and viewing angle. We characterize the
line morphology by its total intensity, the full width at half maximum
(FWHM) and the location of the peak maxima. In order to interpret the
morphological changes in the line we also report the median number of
scatter for each \ly photon in the simulation. For the models where
dust is included we measure the escape fraction as a function of
rotational velocity.

\subsection{Line Morphology}
\label{sec:angles}

The first column in both Figures \ref{fig:CentralSpec} and
\ref{fig:HomSpec} shows that for the static sphere the line properties
are independent of $|\mu|$, as it is expected due
to the spherical symmetry. However, for increasing rotational
velocities, at a fixed optical depth, there are clear signs that this
symmetry is broken.

If the viewing angle is aligned witht the rotation axis, $|\mu|\sim
1$, the \ly line keeps a double peak with some changes as a function
of the rotational velocity that are easier to spot in models where the
\ly photons are homogeneously distributen than in the case of central
emission.

However, for a line of sight perpendicular to the rotation
axis, $|\mu|\sim 0$, the impact of rotation is larger.  This is clear
in  Figure \ref{fig:differentvelocities} where we present the
different line morphologies for $|\mu|\sim 0$ for both the homogeneous
and central configurations. The panels have the same distribution as
Figures \ref{fig:CentralSpec} and \ref{fig:HomSpec}.

There are three clear effects on the line's morphology as the
rotational velocity increases. First, the line broadens; second, the
double peaks reduce their intensity; and third, the intensity at the
line center rises. The last two effects merge to give the
impression that the double peaks are merger into one at high rotational
velocities, a result that is evident for the homogeneously distributed
sources as shown in the dashed lines of Figure
\ref{fig:differentvelocities}.

In

\subsection{Integrated Line Intensity}

We consider possible variations in the integrated flux with respect to
$\theta$.

We define the flux by:
\begin{equation}
F(\mu) = \frac{2\Delta N}{N\Delta \mu}, 
\end{equation}

@@

We

\begin{figure}
\begin{center}
  \includegraphics[width=0.4\textwidth]{muhisto.eps}
\end{center}
\caption{Integrated flux distribution, $F(\mu)$, as a function of the
  viewing angle as parameterized by $\mu$. We show the results for the
  homogeneous source distribution in models with rotational velocities
  $0,100,200,300$\kms and  optical depths $10^5$, $10^6$, $10^7$. We
  see that the distribution is flat, meaning that the flux for all
  viewing angles is the same. The results for the central sources are
  similar. 
\label{fig:muhisto}}
\end{figure}

\subsection{Full Width at Half Maximum}
\label{sec:widthpeak}

@@

@@

\subsection{Peaks Positions}
\label{sec:peaks}

@@

\subsection{Average Number of Scatterings}
\label{sec:scatterings}

@@

@@

Gas bulk rotation has a noticeable effect on the morphology of the
 \ly line. In this section we discuss the implications of these
 findings for the interpretation of available observational data.

@@

\end{itemize}

Comparing our results with recent observed LAEs we find that many 
morphological features such as high central line flux, single peak
profiles  could be explained by gas bulk  rotation present in these
LAEs.
